Parsing Postfix log files

### 2008-03-04 doc/logparser.tex
More corrections after proofreading.

    \item [framework] Is responsible for loading rules, managing data
        structures so actions can retrieve them, finding the rule which
        matches each log line, invoking the correct action, etc.  The
        framework provides the structure which actions and rules plug into.
        The framework is not described in detail in this document because
        it is not particularly interesting.

@@

Where the rules are quite simple and each rule is completely independent of
its fellows, the algorithm is significantly more complicated and highly
internally interdependent.  The algorithm deals with all the complications
of parsing, the eccentricities and oddities of Postfix logs, and presents
the resulting data in a normalised, easy to use representation.  The
algorithm's task is to follow the journey each mail takes through Postfix,
piecing the data extracted by rules into a coherent whole, saving it in a
useful and consistent form, and performing housekeeping duties.

@@

    \item A flow chart showing common paths a mail can take through the
        algorithm, and discussing the emergent behaviour observed in the
        parser.

    \item An explanation of the paths shown in the flow chart.

@@

    \item Additional complications which have arisen during the development
        of this parser are documented, including their solutions and where
        those solutions are implemented in the algorithm

@@

A high level view of the algorithm could be expressed as:

@@

    \item Mail enters the system via \SMTP{} or local submission; and new
        data structure is created for it.

@@

Unfortunately that ignores the many complications encountered.

@@

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by \pid{} for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see \sectionref{timeouts-during-data-phase} for details.  However
in general there isn't any ambiguity about which data structure should be
used for a given log line.}  The original connection will be discarded
unsaved when the client disconnects if it doesn't have any data worth
saving, i.e.\ no rejections.  One unsolved difficulty is distinguishing
between different groups of rejections, e.g.\ when dealing with the
following sequence:

@@

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this occurs
a child mail will be injected into the Postfix queue, but without the
explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
Thus the source is not immediately discernible from the log line in which
the mail first appears; from a strictly chronological reading of the logs
it \textit{usually\/} appears as if the child mail has appeared from thin
air.  Subsequently the parent mail will log the creation of the child mail:

@@

Unfortunately while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition the first log line belonging to the child mail (the log line cited
above properly belongs to the parent mail) is logged by \daemon{qmgr}, so
the order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is
quite paranoid about mail delivery, an excellent characteristic for an
\MTA{} to possess, so it won't log that the child has been created until it
is absolutely certain that the mail has been written to disk.}

@@

Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
image part 2} shows the paths the data representing a mail/connection can
take through the parser algorithm.  The flow chart covers the most common
paths only; there are additional, uncommon paths which are excluded for the
sake of clarity, but are described in \sectionref{additional
complications}.

The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
which is far more complicated than the behaviour described by the
algorithm.  The top-level parser could be written in pseudo-code like so
(with indentation level indicating flow control):

@@

Yet despite the brevity of the algorithm above the (simplified) flow chart
below contains:

@@

Several more states and corresponding branches are omitted for the sake of
clarity (these extra states are caused by solving the difficulties
described in \sectionref{additional complications}).

The solution to the \textit{out of order log log lines\/} complication
encountered during development (\sectionref{out of order log lines})
requires that a list of the acceptable combinations of Postfix programs a
mail can pass through; however this does not correspond to the
\textit{actions\/} a mail must pass through.  Before that complication was
overcome, i.e.\ for the majority of the development of this parser, there
was nothing in the parser explicitly encoding the states or paths a mail
could take; the vast majority of the parser has no need for this flow of
data to be specified.  This eases the process of adding new actions
(described in \sectionref{adding new actions}), as there is no requirement
to explicitly insert the action into an algorithmic version of the flow
chart below.

% This is very likely to move around as the flow chart is as big as
% possible, so if any preceding text changes the flow chart may end up on
% the next page, which may not be a bad thing, maybe.  What would be a bad
% thing is the \clearpage seems to cause a blank page sometimes, if the
% image is just slightly too big.

\subsection{Full algorithm}

\label{full-algorithm}

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.  The parser thus requires both a method of mapping log
lines to the correct mail and a method of specifying the action the log
line represents.  The former is achieved by using the \daemon{smtpd}
process id to identify the correct mail during the initial phase, then
switching to the queueid once the mail has been accepted.  The latter uses
the action field of the rule which matched the log line, executing the code
in the function named by the action.

\sectionref{actions-in-detail} explains the actions in substantive detail;
this section omits such detail because it would clutter and confuse the
algorithm description.  The flow chart in \sectionref{flow-chart} should
also be consulted while reading this section.  For the sake of clarity this
description covers only the most common paths through the algorithm, as
including every path would hinder understanding; the more uncommon paths
are caused by the complications described in \sectionref{additional
complications}, and are covered therein.

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
due to forwarding, or internally generated by Postfix.  Local submission is
the simplest case: a queueid is assigned immediately and the sender address
is logged (action: pickup; flowchart:~2).

    \item First there is a connection from the remote client (action:
        connect; flowchart:~1).

    \item This is followed by rejection of sender address, recipient
        addresses, client \IP{} address or hostname, etc. (action:
        rejection; flowchart:~4); acceptance of one or more mails (action:
        clone; flowchart:~5); or some interleaving of both.

    \item The client disconnects (action: disconnect; flowchart:~6).  If
        Postfix has rejected any \SMTP{} commands the data will be saved to
        the database; if not there won't be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
\sectionref{identifying-bounce-notifications} for details).  Bounce
notifications are the primary example of internally generated mails, though
there may be other types.

    \item Postfix will log the sender and recipient addresses separately
        (action: save\_by\_queueid; flowchart:~9).

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~10) ---
        \sectionref{tracking re-injected mail} discusses this in
        detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables.
        Re-injected mails are described in \sectionref{tracking re-injected
        mail}.

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, as
explained in \sectionref{tracking re-injected mail}.

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery (action: mail\_picked\_for\_delivery; flowchart:~3), in
        which case a new data structure will be created.  The mail will be
        flagged as having unknown origins; this flag should be subsequently
        cleared once the origin has been established.  This may also be an
        indicator that the mail is a bounce notification, see
        \sectionref{identifying-bounce-notifications} for details.

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: track; flowchart:~10).\footnote{Relayed
        delivery is performed by the \SMTP{} client; local delivery means
        local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (described above) and the unknown origin flag will be
        cleared; if not a new data structure will be created.  In both
        cases the new mail is marked as a child of the parent.  The log
        line in question is:

        This second indicator always occurs for re-injected mail but
        typically occurs after the first indicator explained above.  This
        indicator is required to tie the parent and child mails together
        and so is central to the process of tracking re-injected mails.

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent hasn't yet been identified (action: commit;
        flowchart:~15).  Mark the mail as ready for entry in the database
        (flowchart:~16), and wait for the parent to deal with it
        (flowchart:~17).  The mail should not have subsequent log entries;
        only its parent should refer to it.

    \item If the mail is a child mail then it has already been tracked
        (action: commit; flowchart:~18): the data is cleaned up, the child
        is entered in the database (flowchart:~19), then deleted from the
        state tables.  The child mail will be removed from the parent
        mail's list of children (flowchart:~20); if this is the last child
        and the parent has also been entered in the database the parent
        will be deleted from the state tables.

    \item The last alternative is that the mail is a parent mail (action:
        commit; flowchart:~21).  Regardless of the state of its children
        its data is cleaned up and entered in the database (flowchart:~22).
        The parent may have children which are waiting to be entered in the
        database (flowchart:~23); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered, in which case (flowchart:~24) the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~25), and deleted from the state tables.

@@

    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
        creation of bounce messages.  This action creates a new mail if
        necessary; if the mail already exists the unknown origin flag will
        be removed.  The action also marks the mail as a bounce
        notification.  To deal with complication \sectionref{Further out of
        order log lines} this action checks a cache of recent bounce mails
        to avoid creating bogus bounce mails when lines are logged out of
        order.

    \item [CLONE] Multiple mails may be accepted on a single connection, so
        each time a mail is accepted the connection's state table entry
        must be cloned; if the original data structure was used the second
        and subsequent mails would overwrite one another's data.

    \item [COMMIT] Enter the data from the mail into the database. Entry
        will be postponed if the mail is a child waiting to be tracked.
        Once entered, the mail will be deleted from the state tables.

@@

    \item [DELETE] Deals with mail deleted by the administrator using the
        administrative command \daemon{postsuper}.  This action adds a
        dummy recipient address if required, then invokes the COMMIT action
        to handle adding the mail to the database.  The complication this
        action deals with is described fully in \sectionref{Mail deleted
        before delivery is attempted}.

    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
        connection in the database, perform any required cleanup, and
        delete the connection from the state tables.

@@

    \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
        than the local server accepts, the mail will be discarded, and the
        client informed.  See TIMEOUT for further discussion; the two are
        handled in exactly the same way.

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        \sectionref{pickup logging after cleanup}); otherwise it is
        created.  The data from the log line is then saved to the state
        table entry.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See
        \sectionref{timeouts-during-data-phase} for the gory details.

@@

Adding new actions is almost as simple as adding new rules, though actions
are not stored in the database.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action by calling the parser subroutine add\_actions() with the name of the
new action subroutine as a parameter.  No other work is required on the
implementor's part; all of their attention and effort can be focused on the
correctness of their action.  The only negative aspect is that the process
involves editing the parser source code, which makes upgrading to a later
version of the parser more difficult, though by no means impossible.  If
the author of the new action wishes they can take advantage of the parser's
object oriented implementation~\cite{Wikipedia-object-orientation} by
subclassing it and implementing their changes in the derived class,
allowing future upgrading of the parser with greatly reduced chance of
conflicts.\footnote{The real difference between the two approaches is where
the new code is placed.  The simpler option is to change the parser code
directly, but those changes will then have to be made to subsequent
versions of the parser, and as the scope of the changes increases so does
the chance of conflict.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
addition creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the logs used exhibit
the idiosyncrasies and foibles a mail server in the wild must deal with,
but the downside is that significantly altering the configuration just to
log messages from a different Postfix component is not an option.}

@@

The complications described in this section are listed in the order in
which they were encountered during development of the parser.  Each of
these complications caused the parser to operate incorrectly, generating
either warning messages or leaving mails in the state table.  The frequency
of occurrence is much higher at the start of the list, with the first
complication occurring several orders of magnitude more frequently than the
last.  When deciding which problem to address next the most common was
always chosen, as resolving the most common problem would yield the biggest
improvement in the parser, and also prune the greatest number of mails from
the state tables and error messages.

@@

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing.  In the latter case the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format (e.g.\ if it was generated by
        Postfix on another server) the queueid in the message-id will not
        match the queueid of the mail.

@@

If this check is successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \IP{} address, but the database schema
requires many more fields be populated (see \sectionref{connections table}
and \sectionref{results table}).

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
algorithm (\sectionref{timeouts-during-data-phase}).  Sadly Postfix doesn't
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake timeout will be used throughout this
section, but everything applies equally to lost connections.} during the
DATA phase; when this occurs Postfix will discard the mail and the parser
needs to discard the data associated with that mail.  It seems more
intuitive to save the data to the database, but if a timeout occurs there
won't yet have been any data gathered for the mail, so there is none
available to save; the timeout is recorded with the connection data
instead.

@@

    \item If no mails have been accepted yet nothing needs to be done; the
        timeout action ends.  The timeout action is dependant on the clone
        action keeping a list of all mails accepted on each connection.

    \item \ESMTP{} pipelining allows the client to send MAIL FROM, RCPT TO
        and DATA commands in one packet, instead of sending each command
        individually and waiting for the reply before sending subsequent
        commands.  If pipelining is used and Postfix rejects the sender
        address or any of the recipient addresses there is no way for the
        client to tell which command was rejected.  Some clients make no
        attempt to recover and disconnect uncleanly; others presumably
        stop pipelining and restart the mail sending operation.  The client
        could attempt to parse the error message, but that would be
        extremely error prone, whereas not pipelining makes the cause of
        the error message obvious.

@@

During the TIMEOUT action if the mail lacks a \daemon{cleanup} line, the
queueid and other data about the mail, particularly the timestamp of the
last log line, are added to a global cache of timeout queueids.  This cache
is subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action,
and if the queueid is found in the cache and the \daemon{cleanup} line
passes the checks listed above the line will be discarded.

>>>>

    \item [framework] Is responsible for loading rules, managing data
        structures so actions can retrieve them, finding the rule which
        matches each log line, invoking the correct action, etc.  The
        framework provides the structure which actions and rules plug into.
        The framework is not described in detail in this document because
        it is not particularly interesting, but there is some coverage in
        \sectionref{framework}.

@@

Where the rules are quite simple and each rule is completely independent of
its fellows, the algorithm is significantly more complicated and highly
internally interdependent.  The algorithm deals with all the complications
of parsing, the eccentricities and oddities of Postfix logs, presenting the
resulting data in a normalised, easy to use representation.  The
algorithm's task is to follow the journey each mail takes through Postfix,
combining the data extracted by rules into a coherent whole, saving it in a
useful and consistent form, and performing housekeeping duties.

Please refer to \sectionref{parser design} for a discussion of why the
rules, actions and framework have been separated in the parser's design.
In this section algorithm can be taken to mean the combination of framework
and actions.

@@

    \item A flow chart showing common paths a mail can take through the
        algorithm, a discussion of the emergent behaviour observed in the
        parser, and an explanation of the paths shown in the flow chart.

    \item A brief description of the framework.

@@

    \item Additional complications which have arisen during the development
        of this parser are documented, the solutions to those
        complications, and where those solutions are implemented in the
        algorithm

@@

From the viewpoint of an individual mail passing through the parser the
experience could be summarised as:

@@

    \item Mail enters the system via \SMTP{} or local submission.

@@

The framework's high level overview could be expressed as (indentation
denoting flow of control):

Unfortunately both high level views ignore the many complications
encountered.

@@

These complications were encountered early in the parser's implementation
and guided its design and development.

@@

Multiple independent mails may be delivered during one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails won't trample over each other's data.
This must be done every time a mail is accepted, as it's impossible to tell
in advance which connections will accept multiple mails.  Happily once the
mail has been accepted log entries won't be correlated by \pid{} for that
mail any more (its queueid will be used instead), so there isn't any
ambiguity about which mail a given log line belongs
to.\footnote{Unfortunately this statement is not completely accurate: see
\sectionref{timeouts-during-data-phase} for details.  However in general
there isn't any ambiguity about which data structure should be used for a
given log line.}  The original connection will be discarded unsaved when
the client disconnects if it doesn't have any data worth saving, i.e.\ no
rejections.  One unsolved difficulty is distinguishing between different
groups of rejections, e.g.\ when dealing with the following sequence:

@@

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this occurs
a child mail will be injected into the Postfix queue, but without the
explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
Thus the source is not immediately discernible from the log line in which
the mail first appears; from a strictly chronological reading of the logs
it usually appears as if the child mail has appeared from thin air.
Subsequently the parent mail will log the creation of the child mail:

@@

Unfortunately while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition the first log line belonging to the child mail (the log line cited
above belongs to the parent mail) is logged by \daemon{qmgr}, so the order
also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
paranoid about mail delivery, an excellent characteristic for an \MTA{} to
possess, so it won't log that the child has been created until it is
absolutely certain that the mail has been written to disk.}

@@

Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
image part 2} show the paths the data representing a mail can take through
the parser algorithm.  The flow chart covers the most common paths only;
there are additional, uncommon paths which are excluded for the sake of
clarity; details of the deviations can be found in \sectionref{additional
complications}.

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
due to forwarding, or internally generated by Postfix.  Local submission is
the simplest case: a queueid is assigned immediately and the sender address
is logged (action: PICKUP\@; flowchart:~2).

    \item First there is a connection from the remote client (action:
        CONNECT\@; flowchart:~1).

    \item This is followed by rejection of sender address, recipient
        addresses, client \IP{} address or hostname, HELO hostname, etc.\
        (action: REJECTION\@; flowchart:~4); acceptance of one or more
        mails (action: CLONE\@; flowchart:~5); or some interleaving of
        both.

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \SMTP{} commands the data will be saved to
        the database; if not there won't be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
\sectionref{identifying-bounce-notifications} for details).  Bounce
notifications are the primary example of internally generated mails, though
there may be other types.\footnote{Postfix may generate mails to the
administrator when it encounters configuration errors, but such mails are
presumably rare.}

    \item Postfix will log the sender and recipient addresses separately
        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
        \sectionref{tracking re-injected mail} discusses this in
        detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: COMMIT\@; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables.
        Re-injected mails are described in \sectionref{tracking re-injected
        mail}.

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator, \textit{unless\/} the mail is either parent
or child of re-injection, as explained in \sectionref{tracking re-injected
mail}.

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
        which case a new data structure will be created for that mail.  The
        mail will be flagged as having unknown origins; this flag should be
        subsequently cleared once the origin has been established.  This
        may also be an indicator that the mail is a bounce notification,
        see \sectionref{identifying-bounce-notifications} for details.

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
        delivery is performed by the \SMTP{} client; local delivery means
        local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (described above) and the unknown origin flag will be
        cleared; if not a new data structure will be created.  In both
        cases the re-injected mail is marked as a child of the original
        mail.  The log line in question is:

        This second indicator always occurs for re-injected mail but
        typically occurs after the first indicator explained above.  This
        is required to tie the parent and child mails together and so is
        central to the process of tracking re-injected mails.

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent hasn't yet been identified (action: COMMIT\@;
        flowchart:~15).  Mark the mail as ready for entry in the database
        (flowchart:~16), and wait for the parent to deal with it
        (flowchart:~17).  The mail should not have subsequent log entries;
        only its parent should refer to it.

    \item If the mail is a child mail then it has already been tracked
        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
        is cleaned up, the child is entered in the database
        (flowchart:~19), then deleted from the state tables.  The child
        mail will be removed from the parent mail's list of children
        (flowchart:~20); if this is the last child and the parent has also
        been entered in the database the parent will be deleted from the
        state tables.

    \item The last alternative is that the mail is a parent mail (action:
        COMMIT\@; flowchart:~21).  Regardless of the state of its children
        its data is cleaned up and entered in the database (flowchart:~22).
        The parent may have children which are waiting to be entered in the
        database (flowchart:~23); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered (flowchart:~24), in which case the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~25), and deleted from the state tables.

\subsubsection{Emergent behaviour}

The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
which is far more complicated than the behaviour encoded in the algorithm.
The top-level parser could be written in pseudo-code like so (with
indentation level indicating flow control):

@@

Yet despite the brevity of the algorithm above the (simplified) flow chart
in \sectionref{flow-chart} contains:

@@

Several more states and corresponding branches are omitted for the sake of
clarity (these extra states are caused by solving the complications
described in \sectionref{additional complications}).

The solution to the complication in \sectionref{out of order log lines}
(out of order log lines) requires a list of the acceptable combinations of
Postfix programs a mail can pass through; however this does not correspond
to the \textit{actions\/} a mail must pass through.  Before that
complication was overcome, i.e.\ for the majority of the development of
this parser, there was nothing in the parser explicitly encoding the states
or paths a mail could take; no other component in the parser has any need
for this flow of data to be specified.  This eases the process of adding
new actions (described in \sectionref{adding new actions}), as there is no
requirement to explicitly insert the action into an algorithmic version of
the flow chart.

\subsection{Framework}

\label{framework}

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.

This functionality is provided by the framework, which both drives parsing
overall and provides services to the actions it invokes.  The framework is
responsible for loading rules from the database and sanity checking them,
reading log files, matching each rule against the current log line,
invoking the correct action, maintaining state tables, loading and saving
state, displaying a progress indicator, and miscellaneous other tasks.
Actions use services provided by the framework, including storing and
retrieving data in the state tables, extracting and saving data captured by
rules, storage of global data, debugging functions, preparing a mail for
entry in the database, and entering mails in the database.

There is a similarity between this design and the event-driven programming
paradigm commonly used in GUI programs, where one part of the program
responds to events (mouse clicks in a GUI program, log lines being matched
in the parser) and invokes the correct action.

@@

    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
        bounce messages.  This action creates a new mail if necessary; if
        the mail already exists the unknown origin flag will be removed.
        The action also marks the mail as a bounce notification.  To deal
        with complication \sectionref{Further out of order log lines} this
        action checks a cache of recent bounce mails to avoid creating
        bogus bounce mails when lines are logged out of order.

    \item [CLONE] Multiple mails may be accepted on a single connection, so
        each time a mail is accepted the connection's state table entry
        must be cloned and saved in the state tables under its queueid; if
        the original data structure was used then second and subsequent
        mails would overwrite one another's data.

    \item [COMMIT] Enter the data from the mail into the database. Entry
        will be postponed if the mail is a child waiting to be tracked.
        Once entered, the mail will be deleted from the state tables.
        Deletion will be postponed if the mail is the parent of re-injected
        mail.

@@

    \item [DELETE] Deals with mail deleted using Postfix's administrative
        command \daemon{postsuper}.  This action adds a dummy recipient
        address if required, then invokes the COMMIT action to handle
        adding the mail to the database.  The complication this action
        deals with is described fully in \sectionref{Mail deleted before
        delivery is attempted}.

    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
        connection in the database, perform any required cleanup, and
        delete the connection from the state tables.  This action deals
        with aborted delivery attempts
        (\sectionref{aborted-delivery-attempts}).

@@

    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
        than the local server accepts, the mail will be discarded, and the
        client informed.  See TIMEOUT for further discussion; the two are
        handled in exactly the same way.

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        \sectionref{pickup logging after cleanup}); otherwise it is
        created.  The data extracted from the log line is then saved to the
        state table entry.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there isn't.  See
        \sectionref{timeouts-during-data-phase} for the gory details.

@@

Adding new actions is not as simple as adding new rules, though care has
been taken in the parser's design and implementation to make adding new
actions as painless as possible.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action\footnote{The new action must be registered before the rules are
loaded, as it is an error for a rule to reference an unregistered action;
this helps catch mistakes made when adding new rules.} by calling the
framework subroutine add\_actions() with the name of the new action
subroutine as a parameter.  No other work is required on the implementor's
part to integrate the action into the parser; all of their attention and
effort can be focused on the correctness of their action.  The only
negative aspect is that the process involves editing the parser source
code, which makes upgrading to a later version of the parser more
difficult, though by no means impossible.  If the author of the new action
wishes they can take advantage of the parser's object oriented
implementation~\cite{Wikipedia-object-orientation} by subclassing it and
implementing their changes in the derived class, allowing future upgrading
of the parser with greatly reduced chance of conflicts.\footnote{The real
difference between the two approaches is where the new code is placed.  The
simpler option is to change the parser code directly, but those changes
will then have to be made to subsequent versions of the parser, and as the
scope of the changes increases so does the chance of conflict, or mistakes
when copying the action.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
addition creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the logs used exhibit
the idiosyncrasies and peculiarities a mail server in the wild must deal
with, but the downside is that significantly altering the configuration
just to log messages from a different Postfix component is not an option.}

@@

The complications described in this section are listed in the order in
which they were encountered during development of the parser.  Each of
these complications caused the parser to operate incorrectly, generating
either warning messages or leaving mails in the state table.  The frequency
of occurrence is much higher at the start of the list, with the first
complication occurring several orders of magnitude more frequently than the
last.  When deciding which problem to address next the most common was
always chosen, as resolving the most common problem would yield the biggest
improvement in the parser, prune the greatest number of mails from
the state tables and error messages, and make the remaining problems more
apparent.

@@

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing.  In the latter case the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format (e.g.\ if it was generated by
        Postfix on this or another server) the queueid in the message-id
        will not match the queueid of the mail.

@@

If this check is successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \IP{} address, but the database schema
requires many more fields be populated (see \sectionref{connections table}
and \sectionref{results table}).  This heuristic is weak, but thus far it
has not caused any difficulties, and there does not seem to be another
check available to extend the heuristics with.

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser (\sectionref{timeouts-during-data-phase}).  Sadly Postfix doesn't
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake \textit{timeout\/} will be used
throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the data to the database, but if a
timeout occurs there is no data available to save; the timeout is recorded
with the connection data instead.

@@

    \item If no mails have been accepted yet nothing needs to be done; the
        TIMEOUT action ends.  The TIMEOUT action is dependant on the CLONE
        action keeping a list of all mails accepted on each connection.

    \item \ESMTP{} pipelining allows the client to send MAIL FROM, RCPT TO
        and DATA commands in one packet, instead of sending each command
        individually and waiting for the reply before sending subsequent
        commands.  If pipelining is used and Postfix rejects any command
        there is no way for the client to tell which command was rejected.
        Some clients make no attempt to recover and disconnect uncleanly;
        others presumably stop pipelining and restart the mail sending
        operation.  The client could attempt to parse the error message,
        but that would be extremely error prone, whereas not pipelining
        makes the cause of the error message obvious.

        XXX CHECK IF THE DESCRIPTION ABOVE IS CORRECT\@.

@@
